---
title: "Activity - human steps"
author: "Igaal Perez"
date: "22 juillet 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reproducible Research: Peer Assessment 1
================================================================================

For this project, we will process human activity data recorded on a cellular phone. 

##1. Loading and preprocessing the data
First we will load all the required libraries. We will read the file activity.csv
Ensure you modify your working directory.
```{r reading the file , warning=FALSE, messages=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(lattice)

setwd("C:/Users/IP/Documents")
activity<-read.table("activity.csv", sep=",",header=TRUE,  na.strings = "NA", 
quote="'", stringsAsFactors = FALSE)

```
Let us have a look at the dataset "activity". Column names and values are displayed
with extra double quotes.

```{r having a look at the dataset}
head(activity)
```

Let us rename the column names and arrange the values by removing double quotes
and setting the right formats.

```{r Process data}
names(activity)<-c("steps", "date", "interval")

#removing quotes from values
activity$steps<-gsub('"', "",activity$steps)
activity$date<-gsub('"',"", activity$date)
activity$interval<-gsub('"', "",activity$interval)

#converting the dates to Date class
activity$date<-as.Date(activity$date, "%Y-%m-%d")

#converting the steps and  intyerval into numeric
activity$steps<-as.numeric(activity$steps)
activity$interval<-as.numeric(activity$interval)

```
Now the dataset look tidy.
```{r dataset is tidy}
head(activity)
```
  
  

##2. What is mean total number of steps taken per day?
Let us have a look on the histogram of total daily steps. 
```{r histogram of total daimy steps}
# Plotting histogram of total daily steps
byday<-group_by(activity, date)
activitybyday<-summarise(byday, dailysteps=sum(steps))
hist(activitybyday$dailysteps, breaks=40, main="Daily steps histogram",
     xlab="number of daily steps", col="dark green")

```
  
  


##3. What is the average daily activity pattern?
```{r mean and median activity}
activitymeanmedian<-summarise(byday,meansteps=mean(steps, na.rm=TRUE), 
                              mediansteps=median(steps, na.rm=TRUE))
print(activitymeanmedian)
```

Let us now plot the average number of steps by interval:
```{r plotting the average number of steps by interval}
byinterval<-group_by(activity, interval)
activitybyinterval<-summarise(byinterval, averagesteps=mean(steps, na.rm=TRUE))
plot(activitybyinterval$interval, activitybyinterval$averagesteps, type="h", main="Average number of steps by interval"
     , xlab="Interval", ylab="Average number of steps", col="dark blue")
```

We will also find the interval on which the maximum number of steps in average occurs.
```{r interval on which maximum number of steps occurs}
print(paste("The maximum number of steps occurs at interval", 
      activitybyinterval$interval[which.max(activitybyinterval$averagesteps)], 
      "on average across all days."))
```

  
  
##4. Imputing missing values
First let us check how many values are missing:
```{r missing values}
missing<-is.na(activity$steps)
table(missing)
print("There are 2304 missing values in the activity dataset.")
```

Then, let us create a new dataset whith imputed missing values: NAs are replaced
by the mean valuee for the same interval. We will create and use the dataset "activityfilled".
```{r imputing missing values}
activityfilled<-activity 
fillingbyinterval<-function(x, intervalofx){
    if (is.na(x) | is.nan(x))   {
     x<-activitybyinterval$averagesteps[activitybyinterval$interval==intervalofx]   
    } else {
        x=x
    }
}

for(i in 1:17568) {activityfilled$steps[i]<-fillingbyinterval(activityfilled$steps[i], activityfilled$interval[i])}

head(activityfilled)
```
We can compare the "activityfilled" dataset to the "activity"" dataset and check there are no more missing values:
```{r missing values in activityfilled}
missingfilled<-is.na(activityfilled$steps)
table(missingfilled)
```

Now we can plot the histogram of daily total steps when all missing values are replaced.
```{r plotting histogram of total daily steps, no missing values}
bydayfilled<-group_by(activityfilled, date)
activityfilledbyday<-summarise(bydayfilled, dailysteps=sum(steps))
hist(activityfilledbyday$dailysteps, breaks=40, main="Daily steps histogram, no missing values",
     xlab="number of daily steps", col="dark blue")
```

We can notice that the maximum frequency that occurs at the range 11000-12000 is now higher than in the original dataset (12 vs. 7).
Let us check the mean and median values once the NAs are replaced:
```{r mean and median values once NAs are removed}
activityfilledmeanmedian<-summarise(bydayfilled,meansteps=mean(steps, na.rm=TRUE), 
                              mediansteps=median(steps, na.rm=TRUE))
print(activityfilledmeanmedian)
```
  
  

##5. Are there differences in activity patterns between weekdays and weekends?
To answer this question, we will create a factor variable to check which day is 
a weekday or a weekend.
```{r wekday function}
typeofday<-function(jour){
    if(tolower(weekdays(jour)) %in% c("samedi", "dimanche")) {
        "weekend"
    } else {
        "weekday"
    }
}
```
We will add a new column "type" in which will be stored the type of weekday according to its date:
```{r type of weekday}
activityfilled<-mutate(activityfilled, type=NULL)
for (j in 1:17568){
    activityfilled$type[j]<-typeofday(activityfilled$date[j])
}
head(activityfilled)
``` 
Let us group by interval and type. Then we plot using lattice system.
```{r weekday vs. weekend number of steps}
filledbyinterval<-group_by(activityfilled, interval, type)
activityfilledbyinterval<-summarise(filledbyinterval, averagesteps=mean(steps, na.rm=TRUE))

xyplot(averagesteps ~ interval | type, data= activityfilledbyinterval, layout=c(1, 2), type="l")

#releasing the graphical device
dev.off()
dev.set(which = 2)

```
**As a conclusion, we can see that on the weekends, the number of steps is greater during the interval 800-950 by around 70 steps  compared to the weekend.**


